---
title: "Signature Analyzer output processing"
author: "Ankit Dutta and JB Alberge"
date: February 16, 2022
output: 
  html_document:
    toc: true
    toc_depth: 2
    df_print: kable
    keep_md: true
---

# Aggregating different runs from signature analyzer

ARD-NMF doesn't seem to have a kind of estimate of matrix components.
Luckily, our runs all have the same SBS-projected solutions, so we will leverage this to estimate standard deviation around the mean.

```{r cache=FALSE, message=FALSE, warning=FALSE}
library(rhdf5)
library(rsample) # for bootstrapping X

source("../2_wgs/0_annotate_samples.R")
source("../0_utils/signature_analyzer_utils.R")

set.seed(414) # because of bootstrapping

cosine <- function(x, y) x %*% y / sqrt(x%*%x * y%*%y)

# lighten(get_palette("Set1", 8), amount = .4)[c(2, 1, 3, 4)]
sbs.pal <- c("SBS15"="#78B1EC", "SBS2"="#FF8485", "SBS8"="#79D776", "SBS9"="#D08BDB")

dir.create("figures")

paired.participants <- paste0("CTF", 
                              c("001", "002", "003", 
                                "012", "013", "015", "016",
                                "017", "018", "019", 
                                "021", "023", "025", 
                                "034", "036", "046", "047"))
```

## Load and summarize X=HW decomposition

```{r}
# h5_object <- "../data/All_to_Prod5_SA_pcawg_COMPOSITE/nmf_output.h5"
h5_object <- "../data/Paper2022_v2_pcawg_COMPOSITE_March9/nmf_output.h5"

# extract H matrix (signature loading in samples)
signatures <- 0:99 %>%
  map(load_Hmat_SA_h5, h5_object=h5_object) %>%
  bind_rows(.id = "RUN")

summarize.sig.weight <- signatures %>%
  group_by(RUN) %>%
  filter(length(unique(Signature))==4) %>% # 78 out of 100
  mutate(SBS=str_extract(Signature, "SBS[0-9]+"), SBS_Number=str_remove(SBS, "SBS")) %>%
  # mutate(SBS=ifelse(SBS=="SBS2", "SBS2+13", SBS)) %>%
  group_by(Tumor_Sample_Barcodes, RUN) %>%
  mutate(SumWeight = sum(Weight)) %>%
  group_by(SBS, SBS_Number, Tumor_Sample_Barcodes) %>%
  summarize(sum=sum(Weight), mean=mean(Weight), sd=sd(Weight), 
            ci_low=mean-2*sd, ci_high=mean+2*sd, 
            sd_low=mean-sd, sd_high=mean+sd,
            frac_mean=mean(Weight/SumWeight), frac_sd=sd(Weight/SumWeight)) %>%
  # mutate(SBS=factor(SBS, levels=paste0("SBS", unique(sort(SBS_Number))))) %>%
  mutate(SBS=factor(SBS, levels=paste0("SBS", c(2, 8, 9, 15)))) %>%
  arrange(rev(SBS)) %>%
  group_by(Tumor_Sample_Barcodes) %>%
  mutate(cummean=cumsum(mean), 
         cum_cilow=pmax(0, cummean-2*sd), cum_cihigh=cummean+2*sd, 
         cum_sdlow=pmax(0, cummean-sd), cum_sdhigh=cummean+sd) %>%
  mutate(fraccummean=cumsum(frac_mean), 
         fraccum_cilow=pmax(0, fraccummean-2*frac_sd), 
         fraccum_sdlow=pmax(0, fraccummean-frac_sd), 
         fraccum_cihigh=pmin(1, fraccummean+2*frac_sd),
         fraccum_sdhigh=pmin(1, fraccummean+frac_sd))

NRUN=100
summarize.sig.weight <- signatures %>%
  mutate(SBS=str_extract(Signature, "SBS[0-9]+")) %>%
  group_by(SBS) %>%
  filter(length(unique(RUN)) > 2) %>% # need at least three runs to compute descent sd. Lowest currently is SBS72 (1), SBS13 replacing SBS2 in 2, and SBS9 14/100 runs
  group_by(RUN, SBS, Tumor_Sample_Barcodes) %>%
  summarise(Weight=sum(Weight, na.rm = TRUE)) %>% # is several signatures match the same SBS
  ungroup() %>%
  complete(Tumor_Sample_Barcodes, RUN, SBS, fill=list(Weight=0)) %>% # does nt work
  mutate(SBS_Number=str_remove(SBS, "SBS")) %>%
  group_by(Tumor_Sample_Barcodes, RUN) %>%
  mutate(SumWeight = sum(Weight)) %>%
  group_by(SBS, SBS_Number, Tumor_Sample_Barcodes) %>%
  summarize(sum=sum(Weight), mean=mean(Weight), sd=sd(Weight), 
            ci_low=mean-2*sd, ci_high=mean+2*sd, 
            sd_low=mean-sd, sd_high=mean+sd,
            frac_mean=mean(Weight/SumWeight), frac_sd=sd(Weight/SumWeight)) %>%
  mutate(SBS=factor(SBS, levels=paste0("SBS", unique(sort(SBS_Number))))) %>%
  # mutate(SBS=factor(SBS, levels=paste0("SBS", c(2, 8, 9, 15)))) %>%
  arrange(rev(SBS)) %>%
  group_by(Tumor_Sample_Barcodes) %>%
  mutate(cummean=cumsum(mean), 
         cum_cilow=pmax(0, cummean-2*sd), cum_cihigh=cummean+2*sd, 
         cum_sdlow=pmax(0, cummean-sd), cum_sdhigh=cummean+sd) %>%
  mutate(fraccummean=cumsum(frac_mean), 
         fraccum_cilow=pmax(0, fraccummean-2*frac_sd), 
         fraccum_sdlow=pmax(0, fraccummean-frac_sd), 
         fraccum_cihigh=pmin(1, fraccummean+2*frac_sd),
         fraccum_sdhigh=pmin(1, fraccummean+frac_sd))

summarize.sig.weight.annot <- summarize.sig.weight %>% 
  inner_join(clinicaldata, by=c("Tumor_Sample_Barcodes"="Tumor_Sample_Barcode")) %>%
  filter(tissue=="CMMCs" & isRef==TRUE)
  # mutate(participant = fct_reorder(factor(participant), mean, sum, .desc = TRUE))

```

## fraction of APOBEC

```{r}
# in our case they are in the same component but will have to sum whenever SBS2 and SBS13 appear separately
# also see CTF014 is timepoint T1 of CTF004

summarize.sig.weight.annot %>%
  filter(SBS%in%c("SBS2", "SBS13") & participant != "CTF014") %>% View()
  ggplot(aes(fct_reorder(participant, frac_mean), frac_mean)) +
  geom_bar(stat="identity") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90))

```

## Visualize output

```{r fig.width=8, fig.height=5}
dir.create("figures", showWarnings = FALSE)

p1 <- ggplot(summarize.sig.weight.annot, aes(fct_reorder(factor(participant), cummean, max, .desc = TRUE), mean, fill=SBS)) + 
  geom_bar(stat="identity") + 
  geom_linerange(aes(ymin=cum_sdlow, ymax=cum_sdhigh), color="#3B3B3B") +
  facet_grid(rows = "tissue", switch = "y") +
  scale_fill_manual(values=sbs.pal) +
  scale_y_continuous(position = "right") +
  labs(x="", y="Counts") +
  theme_bw() +
  theme(panel.grid = element_blank(), strip.background = element_blank(), axis.text.x = element_text(angle = 90, vjust=.5), legend.position = c(0.9, 0.8))
p2 <- ggplot(summarize.sig.weight.annot, aes(fct_reorder(factor(participant), cummean, max, .desc = TRUE), frac_mean, fill=SBS)) +
  geom_bar(stat="identity") + 
  geom_linerange(aes(ymin=fraccum_sdlow, ymax=fraccum_sdhigh), color="#3B3B3B") +
  facet_grid(rows = "tissue", switch = "y") +
  scale_fill_manual(values=sbs.pal) +
  scale_y_continuous(position = "right", labels=scales::percent) +
  labs(x="", y="Counts") +
  theme_bw() +
  theme(panel.grid = element_blank(), strip.background = element_blank(), axis.text.x = element_text(angle = 90, vjust=.5), legend.position = "none")
# p3 <- ggplot(summarize.sig.weight.annot, aes(fct_reorder(factor(participant), cummean, max, .desc = TRUE), frac_mean, fill=SBS)) +
#   geom_bar(stat="identity") + 
#   geom_linerange(aes(ymin=fraccum_sdlow, ymax=fraccum_sdhigh), color="#3B3B3B") +
#   facet_grid(rows = "tissue", switch = "y") +
#   scale_fill_manual(values=get_palette("uchicago", 8)) +
#   scale_y_continuous(position = "right") +
#   labs(x="", y="Counts") +
#   theme_bw() +
#   theme(panel.grid = element_blank(), strip.background = element_blank(), axis.text.x = element_text(angle = 90, vjust=.5), legend.position = "right")
# legend=get_legend(p3)
# ggsave2(legend, filename = "figures/signatures_legend_new.pdf")
full.sig <- plot_grid(p1, p2, byrow = FALSE, ncol = 1, axis = "vh", align = "lrbt")
ggsave2("figures/all_signature_analyzer_output.pdf", full.sig, width = 8, height = 5)
ggsave2("figures/all_signature_analyzer_output.png", full.sig, width = 8, height = 5)
print(full.sig)

```

# Matched patients

## Post proc and visualisatipn

```{r fig.width=8, fig.height=5}

summarize.sig.weight.annot.paired <- summarize.sig.weight %>% 
  inner_join(clinicaldata, by=c("Tumor_Sample_Barcodes"="Tumor_Sample_Barcode")) %>%
  arrange(rev(tissue)) %>%
  filter(participant %in% paired.participants & isRef==TRUE)

p1 <- ggplot(summarize.sig.weight.annot.paired, aes(fct_reorder(factor(participant), cummean, max, .desc = TRUE), mean, fill=SBS)) + 
  geom_bar(stat="identity") + 
  geom_linerange(aes(ymin=cum_sdlow, ymax=cum_sdhigh), color="#3B3B3B") +
  facet_grid(rows = "tissue", switch = "y") +
  scale_fill_manual(values=sbs.pal) +
  scale_y_continuous(position = "right") +
  labs(x="", y="Counts") +
  theme_bw() +
  theme(panel.grid = element_blank(), strip.background = element_blank(), axis.text.x = element_text(angle = 90, vjust=.5), legend.position = c(0.9, 0.8))
p2 <- ggplot(summarize.sig.weight.annot.paired, aes(fct_reorder(factor(participant), cummean, max, .desc = TRUE), frac_mean, fill=SBS)) +
  geom_bar(stat="identity") + 
  geom_linerange(aes(ymin=fraccum_sdlow, ymax=fraccum_sdhigh), color="#3B3B3B") +
  facet_grid(rows = "tissue", switch = "y") +
  scale_fill_manual(values=sbs.pal) +
  scale_y_continuous(position = "right", labels=scales::percent) +
  labs(x="", y="Counts") +
  theme_bw() +
  theme(panel.grid = element_blank(), strip.background = element_blank(), axis.text.x = element_text(angle = 90, vjust=.5), legend.position = "none")
# p3 <- ggplot(summarize.sig.weight.annot.paired, aes(fct_reorder(factor(participant), cummean, max, .desc = TRUE), frac_mean, fill=SBS)) +
#   geom_bar(stat="identity") + 
#   geom_linerange(aes(ymin=fraccum_sdlow, ymax=fraccum_sdhigh), color="#3B3B3B") +
#   facet_grid(rows = "tissue", switch = "y") +
#   scale_fill_manual(values=get_palette("uchicago", 8)) +
#   scale_y_continuous(position = "right") +
#   labs(x="", y="Counts") +
#   theme_bw() +
#   theme(panel.grid = element_blank(), strip.background = element_blank(), axis.text.x = element_text(angle = 90, vjust=.5), legend.position = "right")
# legend=get_legend(p3)
# ggsave2(legend, filename = "figures/signatures_legend_new.pdf")
full.sig <- plot_grid(p1, p2, byrow = TRUE, ncol = 2, axis = "vh", align = "lrbt")
ggsave2("figures/matched_signature_analyzer_output.pdf", full.sig, width = 8, height = 5)
ggsave2("figures/matched_signature_analyzer_output.png", full.sig, width = 8, height = 5)
print(full.sig)

```

# Mutation reproducibility between tissues from the same patient

Here we will take advantage of the X matrix (so, mutations, not signatures) to compute cosine similarity between runs of CMMCs and BMPCs. Also tried directly on signature weight and work as well. Chip suggested using weighted correlation of signature weight with weight = 1/ sigma^2.

```{r fig.width=4, fig.height=4}

# extract X from X=WH
X <- load_Xmat_SA_h5(h5_object = h5_object)

X.annot <- X %>%
  inner_join(clinicaldata, by=c("Pair"="Tumor_Sample_Barcode")) %>%
  filter(participant %in% paired.participants & isRef==TRUE) %>%
  pivot_wider(id_cols = c(participant, Mutation), names_from = tissue, values_from = Weight)
  
X.boot <- X.annot %>% 
  nest(data = -c(participant)) %>% # grouping the species 
  mutate(boots = map(data, 
  ~bootstraps(.x, times = 1000, apparent = FALSE))) %>%
  unnest(boots) %>% 
  mutate(cosine = map_dbl(splits, ~ with(analysis(.), cosine(BMPCs, CMMCs))))

X.cosine.summary <- X.boot %>% 
  group_by(participant) %>%
  summarise(mean=mean(cosine),
            q025=quantile(cosine, c(.025)),
            q975=quantile(cosine, c(.975)))

median.line <- median(X.cosine.summary$mean)
qts <- c(0, .25, .5, .75, 1)
cohort.cosine.summary <- summarize(X.cosine.summary, quantile(mean, qts))
cohort.cosine.summary

cosine.plot <- ggplot(X.cosine.summary, aes(fct_reorder(factor(participant), mean), mean)) + 
  geom_bar(stat="identity", fill="lightgrey") +
  geom_linerange(aes(ymin=q025, ymax=q975)) +
  geom_hline(yintercept = median.line, linetype=2) +
  annotate(geom="text", y=1.02, x=2, label=paste0("Median = ", scales::percent(median.line, )), size=2.5) +
  scale_y_continuous(labels=scales::percent) +
  labs(x="", y="Cosine similarity (mean, 95% CI)\n") +
  theme_bw() +
  theme(panel.grid = element_blank(), axis.text.x = element_text(angle = 90), aspect.ratio = 1)
print(cosine.plot)
ggsave("figures/cosine.plot.png", cosine.plot, width = 4, height = 4)
ggsave("figures/cosine.plot.pdf", cosine.plot, width = 4, height = 4)
```

# Session info

```{r, eval=TRUE, include=TRUE, comment=NA}
sessionInfo()
```
